Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation

### 2020-09-13 ms/main.tex
formatting, remove pstar; replace partner assoc with social assoc

We constructed an additional model in which partner association emerges from the population structure.
We model a non-overlapping population colonizing a 2D grid of size 100-by-100, where each site is inhabited by one individual, similarly to \citet{lewin2020rockpaperscissors}.
Each individual is characterized by its phenotype which is either a cooperator, $A$, or a defector $B$.
In each generation individuals interact with their neighbors, i.e. individuals that inhabit adjacent sites, in a prisoner's dilemma game (\autoref{table:prisoner_payoff}) and in horizontal cultural transmission.
As before, $T_A$ and $T_B$ are the probabilities to transmit the phenotypes $A$ and $B$ during horizontal transmission.
At each generation, each individual is chosen to "initiate" interactions with probability 1/2.
Initiators then interact with a random neighbor for a prisoners’ dilemma game and a random neighbor (with replacement) for horizontal cultural transmission (as both the game and the transmission are symmetrical, the expected number of each of these interactions per individual per generation is 1).
Therefore, the effective partner association $\alpha$ in this model is the probability that the same neighbor is picked for both interactions, or $\alpha=1/m$, where $m$ is the number of neighbors. On an infinite grid, $m=8$, but on a finite grid $m$ can be lower in edge neighborhoods close to the grid border.

@@

\autoref{fig:spatial} shows that the highest cooperation cost to benefit ratio ($c/b$) that permits the evolution of cooperation in simulations of the spatial model is roughly $\alpha \cdot T / (1-T)$, where $T=T_A=T_B$ is the horizontal transmission rate. This agrees with the condition we derived in Corollary~\ref{sec:sym_hor_trans} for a population without population structure. 
This demonstrates that the partner association parameter $\alpha$ can emerge from local interactions between individuals in a structured population.

@@

%%% Figure: coexistence map - without oblique transmission
\begin{figure}[htb]
\centering
\includegraphics{coexistence_without_oblique.pdf}
\caption{\textbf{Stable and unstable coexistence between cooperation and defection without oblique transmission.}
  The curves show the frequency of the cooperative phenotype $A$ among parents in the next generation $\tilde{p}'$ vs. the current generation $\tilde{p}$ (\autoref{eq:nextgen_parents_vertical_only}).
  The dashed black line is $\tilde{p}'=\tilde{p}$.
  The curves and the dashed line intersect at the polymorphic equilibrium $\tilde{p}^*$ (black circle).
  When the curves are above the dashed line, $\tilde'{p}>\tilde{p}$, then $\tilde{p}$ increase.
  When the curves are below the dashed line, $\tilde'{p}<\tilde{p}$, then $\tilde{p}$ decreases.
  The orange curve, for which the polymorphic equilibrium is stable, is given by $T_A = 0.4$, $T_B = 0.9$, $b = 12$, $c=0.35$, and $\alpha = 0.45$, which give $\gamma_2<c<\gamma_1$ (\autoref{eq:cost_boundaries})
  The blue curve, for which the equilibrium is unstable, is given by $T_A = 0.5$, $T_B = 0.1$, $b = 1.3$, $c=0.904$, and $\alpha = 0.4$, which give $\gamma_1<c<\gamma_2$.
  In both cases there is no oblique transmission, $v=1$; see \autoref{fig:coexistence_with_oblique} for $v<1$.
  }
\label{fig:coexistence_without_oblique}
\end{figure}

%%% Figure: Conditions for fixation

\begin{figure}[htb]
  \centering
  % TODO for some reason figs a and b have a lot of white space on their right side

    \includegraphics[scale=0.75]{boundiries_alpha0.pdf}

    \includegraphics[scale=0.75]{boundiries_non_zero_alpha.pdf}

    \includegraphics[scale=0.75]{boundiries_withAlpha_y_axis.pdf}
    \caption{\textbf{Evolution of cooperation under vertical and horizontal cultural transmission.} 
The figure shows the global fixation of cooperation (green), global fixation of defection (red), fixation of either cooperation or defection depending on the initial conditions, i.e. unstable coexistence (blue), and stable coexistence of cooperation and defection (yellow).
In all cases the horizontal bias ($T_A-T_B$) is on the x-axis.
(\textbf{a-b})~the cost of cooperation $c$ is on the y-axis; the cost boundaries $\gamma_1$ and $\gamma_2$ (\autoref{eq:cost_boundaries}) are the solid and dashed lines. 
(\textbf{c})~the social association $\alpha$ is on the y-axis; the social association boundaries $a_1$ and $a_2$ (\autoref{eq:boundries_assortative_meeting}) are the solid and dashed lines.
Here, $b=1.3$, $T_A=0.4$. (\textbf{a})~$\alpha = 0$.(\textbf{b})~$\alpha = 0.7$. (\textbf{c})~$c = 0.35$.
  % TODO missing text descriptions of conditions on (3)
  	}
    \label{fig:results2}
\end{figure}

@@

%%% Figure: Spatial model
\begin{figure}[htb]
  \centering
    \includegraphics[scale=0.7]{spatial.pdf}
  \caption{
  \textbf{Evolution of cooperation in a spatial model.}
  The expected frequency of cooperators in a structured population after 10,000 generations is shown (red for 0\%, green for 100\%) as function of both the ratio between the cost and benefit of cooperation ($c/b$) on the y-axis, and the horizontal transmission rate (without transmission bias, $T=T_A=T_B$) on the x-axis.
  The population evolves on a 100-by-100 grid. Selection, cooperation, and horizontal cultural transmission are all local between adjacent sites.
  The black curve represents the condition for the evolution of cooperation in a well-mixed population with partner association, $c/b<\alpha T / (1-T)$, where $\alpha=1/8$; see \autoref{eq:equal_transmission}.
  Note that in the structured population, selection is local, whereas in the unstructured population, selection is global. This can explain the small difference in the results.
  Here, population size is $10,000$ (100-by-100 grid); cost of cooperation, $c=0.05$. 50 simulations were executed per parameter set. Simulations were stopped at generation 10,000 or if one of the phenotypes fixed.
  }
  \label{fig:spatial}
\end{figure}

>>>>

We constructed an additional model in which social association emerges from the population structure.
We model a non-overlapping population colonizing a 2D grid of size 100-by-100, where each site is inhabited by one individual, similarly to \citet{lewin2020rockpaperscissors}.
Each individual is characterized by its phenotype which is either a cooperator, $A$, or a defector $B$.
In each generation individuals interact with their neighbors, i.e. individuals that inhabit adjacent sites, in a prisoner's dilemma game (\autoref{table:prisoner_payoff}) and in horizontal cultural transmission.
As before, $T_A$ and $T_B$ are the probabilities to transmit the phenotypes $A$ and $B$ during horizontal transmission.
At each generation, each individual is chosen to "initiate" interactions with probability 1/2.
Initiators then interact with a random neighbor for a prisoners’ dilemma game and a random neighbor (with replacement) for horizontal cultural transmission (as both the game and the transmission are symmetrical, the expected number of each of these interactions per individual per generation is 1).
Therefore, the effective social association $\alpha$ in this model is the probability that the same neighbor is picked for both interactions, or $\alpha=1/m$, where $m$ is the number of neighbors. On an infinite grid, $m=8$, but on a finite grid $m$ can be lower in edge neighborhoods close to the grid border.

@@

\autoref{fig:spatial} shows that the highest cooperation cost to benefit ratio ($c/b$) that permits the evolution of cooperation in simulations of the spatial model is roughly $\alpha \cdot T / (1-T)$, where $T=T_A=T_B$ is the horizontal transmission rate. This agrees with the condition we derived in Corollary~\ref{sec:sym_hor_trans} for a population without population structure. 
This demonstrates that the social association parameter $\alpha$ can emerge from local interactions between individuals in a structured population.

@@

%%% Figure: coexistence map - without oblique transmission
\begin{figure}[htb]
\centering
\includegraphics[scale = 0.75]{coexistence_without_oblique.pdf}
\caption{\textbf{Stable and unstable coexistence between cooperation and defection without oblique transmission.}
  The curves show the frequency of the cooperative phenotype $A$ among parents in the next generation $\tilde{p}'$ vs. the current generation $\tilde{p}$ (\autoref{eq:nextgen_parents_vertical_only}).
  The dashed black line is $\tilde{p}'=\tilde{p}$.
  The curves and the dashed line intersect at the polymorphic equilibrium $\tilde{p}^*$ (black circle).
  When the curves are above the dashed line, $\tilde'{p}>\tilde{p}$, then $\tilde{p}$ increase.
  When the curves are below the dashed line, $\tilde'{p}<\tilde{p}$, then $\tilde{p}$ decreases.
  The orange curve, for which the polymorphic equilibrium is stable, is given by $T_A = 0.4$, $T_B = 0.9$, $b = 12$, $c=0.35$, and $\alpha = 0.45$, which give $\gamma_2<c<\gamma_1$ (\autoref{eq:cost_boundaries})
  The blue curve, for which the equilibrium is unstable, is given by $T_A = 0.5$, $T_B = 0.1$, $b = 1.3$, $c=0.904$, and $\alpha = 0.4$, which give $\gamma_1<c<\gamma_2$.
  In both cases there is no oblique transmission, $v=1$; see \autoref{fig:coexistence_with_oblique} for $v<1$.
  }
\label{fig:coexistence_without_oblique}
\end{figure}

%%% Figure: Boundaries figure - only vertical

\begin{figure}[htb]
  \centering       
    \includegraphics[width=0.45\textwidth]{Result2_c_zero_alpha.eps}
	~
    \includegraphics[width=0.45\textwidth]{Result2_c_non_zero_alpha.eps}
    ~
    \includegraphics[width=0.45\textwidth]{Result2_alpha.eps}

    \caption{\textbf{Evolution of cooperation under vertical and horizontal cultural transmission.} 
The figure shows the global fixation of cooperation (green), global fixation of defection (red), fixation of either cooperation or defection depending on the initial conditions, i.e. unstable coexistence (blue), and stable coexistence of cooperation and defection (yellow).
In all cases the horizontal bias ($T_A-T_B$) is on the x-axis.
(\textbf{a-b})~The cost of cooperation $c$ is on the y-axis; the cost boundaries $\gamma_1$ and $\gamma_2$ (\autoref{eq:cost_boundaries}) are the solid and dashed lines. 
(\textbf{c})~social association $\alpha$ is on the y-axis; the social association boundaries $a_1$ and $a_2$ (\autoref{eq:boundries_assortative_meeting}) are the solid and dashed lines.
Here, $b=1.3$, $T_A=0.4$. (\textbf{a})~$\alpha = 0$.(\textbf{b})~$\alpha = 0.7$. (\textbf{c})~$c = 0.35$.
  % TODO missing text descriptions of conditions on (3)
  	}
    \label{fig:results2}
\end{figure}

%%% Figure: Boundaries figure - vertical and oblique

\begin{figure}[htb]
  \centering       
    \includegraphics[width=0.45\textwidth]{Result3_TA_TB.eps}~
    \includegraphics[width=0.45\textwidth]{Result3_TB_TA.eps}
    \caption{\textbf{Evolution of cooperation under vertical, oblique, and horizontal cultural transmission.} 
The figure shows the global fixation of cooperation (green), global fixation of defection (red), fixation of either cooperation or defection depending on the initial conditions, i.e. unstable coexistence (blue), and stable coexistence of cooperation and defection (yellow).
The vertical transmission rate $v$ is on the x-axis.
Horizontal transmission bias (\textbf{a}) for cooperation, $T_A>T_B$ and (\textbf{b}) for defection, $T_A<T_B$.
Cost boundaries $\hat\gamma_1$ and $\hat\gamma_2$ (\autoref{eq:cost_boundaries_v}) are represented by the solid and dashed lines. 
(\textbf{a}) $b=1.2$, $T_A = 0.5$, $T_B = 0.4$, $\alpha = 0.4$.
(\textbf{b}), $b=2$, $T_A=0.5$, $T_B = 0.7$, $\alpha = 0.7$.
	}
    \label{fig:result3}
\end{figure}

@@

%%% Figure: Spatial model
\begin{figure}[htb]
  \centering
    \includegraphics[scale=0.7]{spatial.pdf}
  \caption{
  \textbf{Evolution of cooperation in a spatial model.}
  The expected frequency of cooperators in a structured population after 10,000 generations is shown (red for 0\%, green for 100\%) as function of both the ratio between the cost and benefit of cooperation ($c/b$) on the y-axis, and the horizontal transmission rate (without transmission bias, $T=T_A=T_B$) on the x-axis.
  The population evolves on a 100-by-100 grid. Selection, cooperation, and horizontal cultural transmission are all local between adjacent sites.
  The black curve represents the condition for the evolution of cooperation in a well-mixed population with social association, $c/b<\alpha T / (1-T)$, where $\alpha=1/8$; see \autoref{eq:equal_transmission}.
  Note that in the structured population, selection is local, whereas in the unstructured population, selection is global. This can explain the small difference in the results.
  Here, population size is $10,000$ (100-by-100 grid); cost of cooperation, $c=0.05$. 50 simulations were executed per parameter set. Simulations were stopped at generation 10,000 or if one of the phenotypes fixed.
  }
  \label{fig:spatial}
\end{figure}
